On the Vim editor:\\ vim features and tweaks after 10 years of usage

### 2017-11-04 article/article.tex
added useful Vim src/Makefile. Used for the article.

\section{Introduction}
Vim is a very complex editor,
considered by the Linux community to be
matched only by Emacs.
They both are the standard advanced text editors
of the open source community
and have been developed for decades.
This document describes the Vim editor
and proposes a set of enhancements of the user
experience which
are implemented as Vim Plugins.
The editor has a very mature documentation
and the contents herein presented is a
report on the overall understandings I
have of Vim after a little bit more
of ten years using it.
The purposes of this document are:
\begin{itemize}
  \item to help new users in grasping Vim essentials
  and convenient practices.
  \item To make a sound overview of the editor available.
  \item To record the view of the editor that
  an user (me) has after $10$ years of usage.
  \item To confront my usage with that of other experienced
  users. This is helpful for me, but also for the other users
  as they might benefit from this content and from discussions
  that might arise from it.
\end{itemize}

I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
I used Vim for writting and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
Eventually, I edited database files and other types os files.
With Vim, I mostly write software (web and scientific),
music, poems and short stories.
It is very useful because:
\begin{itemize}
  \item it is meant to be a plain text (e.g. ascii, utf8) editor
  and does not (by standard) insert special charaters (e.g. for formating, with binary instructions).
  \item It has a powerful architecture and set of commands.
  \item It is highly configurable and most often the users
  has a set of commands.
\end{itemize}

@@

\subsection{Using Vim's modes}
Vim has some basic modes of usage:
\begin{itemize}
  \item Normal mode: used for changing
  the position of the cursor or the text displayed
  at the window.
  A code goal of the normal mode it to allow fast
  navigation of the document while allowing
  the typist to maintain its fingers on the home row
  (i.e. on the center of the keyboard).
  The mode is also used for manupulating text
  (e.g. copy, paste, delete, change case).
  \item Insert mode: for inserting text.
  \item Command-line mode: for entering Ex commands.
  \item Ex mode: similar to command-line mode,
  but more specialized for running various Ex commands.
  \item Visual mode: for making, manipulating and navigating
  selections of texts.
  \item Select mode: similar to visual mode but
  favors CUA\footnote{IBM Common User Access: \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
\end{itemize}

@@

For most basic and naive navigation, one should check Section~\ref{minimum}.
Most often, one uses:
\begin{itemize}
  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
and whole page down and up, although these commands might
be set to scroll a different number of lines.
  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
  \item \texttt{(w,b,e)} to move to the next, previous and end-of-next word. There motions iterate over sequences of characters separated
    by special characters (e.g. punctuation and parenthesis) as
    specified by the output of \texttt{:se iskeyword}.
    To interate over space-separated tokens, use \texttt{W,B,E}).
    To move to the end of last word, one might use \texttt{be}
    or \texttt{ge}.
  \item \texttt{(fX,tX)} to move to or just before any X character.
  \item \texttt{\},\{} for moving to the next blank line.
  \item Search with / or ?, although these are in truth command-line 
  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
    jump list.
  \item \texttt{'X,`X} to move the cursor to a mark bond to the alphanumeric character X: \texttt{`X} moves to the exact position while \texttt{'X} moves to
    the first non-blank character of the line.
    A mark is registered by the user in any cursor position
    by typing \texttt{mX}, where X is any letter.
    If X is lowercase, the mark is local to the buffer (the file),
    if it is uppercase or numeric, it is global to the Vim session.
\end{itemize}

@@

\subsubsection{Insertion}
Once in the insertion mode, the character keys
input the characters at the cursor position.
One can exit insert mode by pressing \texttt{<Esc>},
and Vim will be put in normal mode.
Most used commands in insert mode include:
\begin{itemize}
  \item \texttt{Ctrl+o} to execute one and only command in normal mode.
  \item \texttt{asdjn}
\end{itemize}

@@

@@

\subsection{Standard configuration files and directories and My .vim/vimrc}
\subsection{Spell and spelllang (en and pt\_br)}
\subsection{Tabs, splits, buffers and namespaces}
\& \% \$ and the following
\subsection{Mappings and abbreviations}
\subsection{Macros and registers for copy and paste}
\subsection{History of commands}
\subsection{Lists of Vim state variables}
% of markers, jumps, registers, commands (hist)
% :
\subsection{Undo}
\subsection{Scripting, Functions, Vimscript and Python}
\subsection{Plugins}
\subsubsection{Standard features}
\subsubsection{Writting plugins}
\subsubsection{Plugin systems: usage for using and writing plugins}
\subsection{Colour}
\subsubsection{Standard, 8 and 16 bits, and true color, Screen/Byobu}
\subsubsection{Gruvebox, Solarize and other colourschemes}
\subsection{Fonts}
\subsubsection{Cools nd popular fonts}
\subsubsection{How to set fonts in xterm, gnome-terminal and GVim.}
\subsection{Verbosing, logs and possibilities of using it to study your own}
\subsection{age (often used commands and typed sequences).}
\subsection{Highlighting}
\subsection{Bash and Vim commands}
\subsection{Compiling, standard features and plugins}
\subsection{Quickfix}
\subsection{Persistence and vimrc}
% markers, undo, buffers, registers, bookmarked directories,
% vimrc for settings
\section{Issues and plugins}\label{issues}

@@

\appendix
\section{Example of usage session}
% open file, split, tab, find, replace char,
% append to line, visual block, etc.
\section{Key naming in Vim documentation}
% CTRL, <C-X> <C-S-X>
% Special chars..
\section{My vimrc file}
In this Appendix is my vimrc file.
Although it has comments, one should
look for the options that (s)he does
not understand, as a thorough explanation
of the file is tedious and out of the scope
of this document.

\section{Overview of my usage of Vim}
% historical usage, other editors,
% most often edited file types

>>>>

\section{Introduction}
Vim is a very complex editor,
considered by the Linux community to be
matched only by Emacs.
They both are the standard advanced text editors
of the open source community
and have been developed for decades.
This document describes the Vim editor
and proposes a set of enhancements of the user
experience which
are implemented as Vim Plugins.
The editor has a very mature documentation
and the contents herein presented is a
report on the overall understandings I
have of Vim after a little bit more
of ten years using it.
The purposes of this document are:
\begin{itemize}
  \item to help new users in grasping Vim essentials
  and convenient practices.
  \item To make a sound overview of the editor available.
  \item To record the view of the editor that
  an user (me) has after $10$ years of usage.
  \item To confront my usage with that of other experienced
  users. This is helpful for me, but also for the other users
  as they might benefit from this content and from discussions
  that might arise from it.
  \item To propose some enhancements to Vim through simple plugins.
\end{itemize}

Advanced users might just focus on the proposed enhancements and
in some of the basic topics, such as in using numerous terminals 
(it works pretty well!) and using sessions (Section~\ref{sessions}).
These make Vim very convenient e.g. for users that edit multiple files,
run scripts and compile programs and latex files.
I used for years multiple terminal abs and windows,
and sometimes a byobu/screen terminal multiplexer.
Nowadays I am using standard Vim windows as terminals
such as described in Section~\ref{terminals}.

I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
I used Vim for writing and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
Eventually, I edited database files and other types os files.
With Vim, I mostly write software (web and scientific),
music, poems and short stories.
It is very useful because:
\begin{itemize}
  \item it is meant to be a plain text (e.g. ascii, utf8) editor
  and does not (by standard) insert special charaters (e.g. for formating, with binary instructions).
  \item It has a powerful architecture and set of commands.
  \item It is highly configurable and most often the users
  has a set of commands.
\end{itemize}

@@

\subsection{Using Vim's modes}
Vim has some basic and fully implemented modes of usage:
\begin{itemize}
  \item Normal mode: used for changing
  the position of the cursor or the text displayed
  at the window.
  A code goal of the normal mode it to allow fast
  navigation of the document while allowing
  the typist to maintain its fingers on the home row
  (i.e. on the center of the keyboard).
  The mode is also used for manupulating text
  (e.g. copy, paste, delete, change case).
  \item Insert mode: for inserting text.
  \item Command-line mode: for entering Ex commands.
  \item Ex mode: similar to command-line mode,
  but more specialized for running various Ex commands.
  \item Visual mode: for making, manipulating and navigating
  selections of texts.
  \item Select mode: similar to visual mode but
  favors CUA\footnote{IBM Common User Access: \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
\end{itemize}

There is another basic mode, but it is not fully implemented:
the Terminal-Job mode.

@@

For most basic and naive navigation, one should check Section~\ref{minimum}.
Most often, one uses:
\begin{itemize}
  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
and whole page down and up, although these commands might
be set to scroll a different number of lines.
  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
  \item \texttt{(w,b,e)} to move to the next, previous and end-of-next word. There motions iterate over sequences of characters separated
    by special characters (e.g. punctuation and parenthesis) as
    specified by the output of \texttt{:se iskeyword}.
    To interate over space-separated tokens, use \texttt{W,B,E}).
    To move to the end of last word, one might bumba use \texttt{be}
    or \texttt{ge}.
  \item \texttt{(fX,tX)} to move to or just before any X character.
  \item \texttt{\},\{} for moving to the next blank line.
  \item Search with / or ?, although these are in truth command-line 
  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
    jump list.
  \item \texttt{'X,`X} to move the cursor to a mark bond to the alphanumeric character X: \texttt{`X} moves to the exact position while \texttt{'X} moves to
    the first non-blank character of the line.
    A mark is registered by the user in any cursor position
    by typing \texttt{mX}, where X is any letter.
    If X is lowercase, the mark is local to the buffer (the file),
    if it is uppercase or numeric, it is global to the Vim session.
\end{itemize}

@@

\subsubsection{Insertion}
Once in the insertion mode, the character keys
input the characters at the cursor position.
One can exit insert mode by pressing \texttt{<Esc>},
and Vim will be put in normal mode.
Most used commands in insert mode include:
\begin{itemize}
  \item \texttt{Ctrl+o} to execute one and only command in normal mode.
	  This enter a secondary mode (see Section~\ref{modes})
  \item \texttt{asdjn}
\end{itemize}

@@

\subsubsection{Terminal-Job mode}\label{terminal}

@@

\subsection{Standard configuration files and directories and My .vim/vimrc}
\subsection{Spell and spelllang (en and pt\_br)}
\subsection{Tabs, splits, buffers and namespaces}
\& \% \$ and the following
\subsection{Mappings and abbreviations}
\subsection{Macros and registers for copy and paste}
\subsection{History of commands}
\subsection{Lists of Vim state variables}
% of markers, jumps, registers, commands (hist)
% :
\subsection{Undo}
\subsection{Scripting, Functions, Vimscript and Python}
\subsection{Plugins}
\subsubsection{Standard features}
\subsubsection{Writting plugins}
\subsubsection{Plugin systems: usage for using and writing plugins}
\subsection{Colour}
\subsubsection{Standard, 8 and 16 bits, and true color, Screen/Byobu}
\subsubsection{Gruvebox, Solarize and other colourschemes}
\subsection{Fonts}
\subsubsection{Cools nd popular fonts}
\subsubsection{How to set fonts in xterm, gnome-terminal and GVim.}
\subsection{Verbosing, logs and possibilities of using it to study your own}
\subsection{age (often used commands and typed sequences).}
\subsection{Highlighting}
\subsection{Bash and Vim commands}
\subsection{Compiling, standard features and plugins}
\subsection{Quickfix}
\subsection{Persistence and vimrc}
\subsubsection{Sessions}\label{sessions}
% markers, undo, buffers, registers, bookmarked directories,
% vimrc for settings
\section{Issues and plugins}\label{issues}
\subsection{Sessions}
\subsection{AA messages}
\subsection{Slick Vim}
% tab line and status line and ruler

@@

\appendix
\section{Example of usage session}
% open file, split, tab, find, replace char,
% append to line, visual block, etc.
\section{Key naming in Vim documentation}
% CTRL, <C-X> <C-S-X>
% Special chars..
\section{My vimrc file and usage}
In this Appendix is my vimrc file.
Although it has comments, one should
look for the options that (s)he does
not understand, as a thorough explanation
of the file is tedious and out of the scope
of this document.

Mostly, I toggle the status line with \texttt{\\\\B}
and the tab line with \texttt{\\\\T}.
Save and close windows with \texttt{\\w} and \texttt{\\q}.

\section{Overview of my usage of Vim}
% sessions, terminal for ipython3 and evince
% alternative with byobu or many terminals
% historical usage, other editors,
% most often edited file types
